Windowed least-squares model reduction for dynamical systems

### 2019-10-06 main.tex
done for day

%\title{Model Reduction via \methodName}
%\title{Windowed least--squares reduced-order models for dynamical systems}
\title{Windowed least-squares model reduction for dynamical systems}

@@

The \methodAcronym\ approach can be viewed as a hybrid space--time method and displays 
commonalities with several related efforts. 
First, Ref.~\cite{bui_thesis} tersely formulated a space--time least-squares ROM
and connected this formulation with optimal control; specifically it mentioned the
optimize-then-discretize vs.\ discretize-then-optimize approaches. This work
did not fully develop this approach, eschewing it 
for 
sequential residual minimization in time (i.e., LSPG). \KTC{HERE} The present work thus
formally develops (and extends) several of the concepts put forth
in~\cite{bui_thesis}. Next, Ref.~\cite{constantine_strom} developed a
space--time residual minimization formulation for model interpolation.  The
present work distinguishes itself from Ref.~\cite{constantine_strom} in that
(1) in addition to trial subspaces obtained via space--time projection, we
consider trial subspaces obtained via spatial projection only (i.e., a classic
spatial basis is employed, opposed to a space--time basis)% -- in which case
%the generalized coordinates comprise a vector of functions
, and (2) we
minimize the time-continuous residual over space--time
windows;~\cite{constantine_strom} considers the solution of a fully discrete
optimization problem over the entire space--time domain. It is noted that,
similar to the current work, a time-continuous residual-minimization
formulation is used as a starting point in~\cite{constantine_strom}, and as
such thematic similarities exist between~\cite{constantine_strom} and the
present work. Lastly, Ref.~\cite{choi_stlspg} develops a space--time
formulation of the least-squares Petrov--Galerkin approach. The formulation
minimizes the \textit{fully discrete} residual over the entire space--time
domain within a given space--time trial subspace. 
The present work distinguishes itself from Ref.~\cite{choi_stlspg} in that
(1) we perform model reduction of the time-continuous system, (2) we
consider trial subspaces obtained via both spatial projection only and space--time projection,
 and (3) residual minimization is performed over time
windows; Ref.~\cite{choi_stlspg} considers trial subspaces obtained via space--time projection and minimizes the
fully-discrete residual over the entire time domain.

\begin{figure} 
\begin{centering} 
\includegraphics[trim={0.1cm 0cm 1cm 0cm},clip,width=0.99\textwidth]{diagram.pdf} 
\caption{Relationship diagram for the \methodAcronym\ framework for model
	reduction.} 
\label{fig:flowchart} 
\end{centering} 
\end{figure}

 The novel contributions and findings of the present work are:
\begin{enumerate}
%\item The development of a time-continuous residual minimization framework for reduced-order models.
\item The development of the \methodNameLower\ (\methodAcronym) approach for dynamical systems. The approach sequentially minimizes the time-continuous full-order model residual over time windows.
\item The derivation of the stationary conditions for \methodAcronymROMs\ leveraging spatial and space--time dimension reduction trial subspaces. The case of spatial dimension reduction trial subspaces is of particular interest: the stationary conditions are derived via the Euler--Lagrange equations and comprise a coupled 
two-point Hamiltonian boundary value problem containing a forward and backward system. The forward system, which is forced by an auxiliary costate, evolves the generalized coordinates of the ROM. The backwards system, which is forced by the FOM residual evaluated about the ROM state, governs the dynamics of the costate. 
\item An exploration of the use of both direct (discretize then optimize) and indirect (optimize then discretize) solution techniques. Conditions under which \methodAcronym\ recovers Galerkin, LSPG, and space--time LSPG are presented.
\item Error analysis of the \methodAcronym\ approach for trial subspaces with spatial dimension reduction. In particular, it is shown that the error in \methodAcronymROMs\ is bounded \textit{a priori} by a combination of the error at the start of the window and the integrated FOM ODE residual evaluated at the FOM state projected onto the trial subspace. 
\item Numerical experiments for trial subspaces with spatial dimension reduction demonstrating two key findings:
\begin{itemize}
\item Residual minimization over larger time windows leads to more stable and physically relevant solutions as well as lower space--time residuals. 
\item Minimizing the residual over a larger time window does not necessarily lead to a more accurate solution in the $\elltwo$ norm. Residual 
minimization over an intermediary-sized time-window leads to the lowest error in the $\elltwo$ norm.
\end{itemize}
\end{enumerate}

The paper proceeds as follows: Section~\ref{sec:math}
outlines the mathematical setting for the full-order model, along with the Galerkin, LSPG, and
space--time LSPG ROMs; Section~\ref{sec:tclspg} outlines the proposed \methodAcronym\
formulation; Section~\ref{sec:numerical_techniques}
outlines numerical techniques for solving \methodAcronymROMs, including both direct and
indirect methods; Section~\ref{sec:analysis} provides
equivalence conditions and error analysis for \methodAcronymROMs;
Section~\ref{sec:numerical_experiments} presents numerical experiments;
Section~\ref{sec:conclude} provides conclusions and perspectives.
We denote vector-valued functions with stylized bold text (i.e., $\boldsymbol
x$), vectors with standard bold text (i.e., $\mathbf{x}$), 
matrices with capital bold letters (i.e., $\mathbf{X} \equiv \begin{bmatrix}
\mathbf{x}_1 & \cdots & \mathbf{x}_r\end{bmatrix}$), and spaces with
calligraphic fonts (i.e., $\mathcal{X}$).
%Model reduction of nonlinear dynamical systems is a growing research area.
%The least-squares Petrov-Galerkin (LSPG) approach has gained popularity as a model reduction technique for nonlinear dynamical systems. Derived at the fully discrete level, the LSPG approach solves a least-squares problem that minimizes the fully discrete residual of the reduced-order model at each time-step. LSPG has been shown to be an effective formulation for nonlinear model reduction on complex problems of interest.
%\begin{comment} 
\section{Mathematical setting}\label{sec:math}
We consider the system of nonlinear ordinary differential equations (ODEs),
\begin{equation}\label{eq:FOM}
 \stateFOMDotArg{t}  = \velocity(\stateFOMArg{}{t},t), \qquad \stateFOM(0) =
	\stateFOMIC,\qquad t \in [0,T]
\end{equation}
where $\stateFOMDot \equiv \frac{d \stateFOM}{dt}$, 
$T \in \mathbb{R}^+$ denotes the final time, $\stateFOM \in \RR{\fomdim}
	\otimes \timeSpace $ with $\stateFOM: [0,T] \rightarrow  \RR{\fomdim}$
	denotes the state implicitly defined as the solution to initial value
	problem~\eqref{eq:FOM}, 
	$\timeSpace$ denotes the set of real-valued functions acting on the time
	domain, 
 $\stateFOMIC \in \mathbb{R}^{\fomdim}$ denotes the initial condition,
	and $\velocity: \mathbb{R}^{\fomdim} \times [0,T] \rightarrow
	\mathbb{R}^{\fomdim}$ with $(\stateyDiscrete,t) \mapsto
	\velocity(\stateyDiscrete,t)$ denotes the velocity, which is possibly
	nonlinear in its first argument. We refer to Eq.~\eqref{eq:FOM} 
	as the ``full-order model" (FOM) ODE.

@@

\subsection{Projection-based reduced-order models}
Projection-based ROMs generate approximate solutions to the FOM
	ODE~\eqref{eq:FOM} by approximating the state in a low-dimensional trial
	subspace. For dynamical 
	systems, two distinct types of space--time trial subspaces are commonly used: 
\begin{enumerate} 
	\item \textit{Space–time trial subspaces with spatial dimension reduction (\spatialAcronym)}. These 
	trial subspaces reduce only the spatial dimension of the FOM, and are commonly used
		in model-reduction methods based on spatial projection followed by time
		integration (e.g., Galerkin projection, LSPG projection.)
	\item \textit{Space–time trial subspaces with spatial and temporal dimension reduction (\spaceTimeAcronym)}. These 
	trial subspaces reduce both the spatial and temporal dimensions of the FOM,
		and are commonly employed 
		employed in model-reduction methods based on space--time projection (e.g.,
		space--time LSPG projection).
\end{enumerate}
 We now describe these two types of trial subspaces and their application to 
	Galerkin, LSPG, and space--time LSPG projection. 
	\KTC{HERE}

@@

\section{Acknowledgments}
The authors thank Yukiko Shimizo and Patrick Blonigan for numerous conversations from which this work benefited. 
E.\ Parish acknowledges an appointment to the Sandia National Laboratories'
John von Neumann Postdoctoral Research Fellowship in Computational Science. This paper
describes objective technical results and analysis. Any subjective views or
opinions that might be expressed in the paper do not necessarily represent the
views of the U.S.\ Department of Energy or the United States Government.
Sandia National Laboratories is a multimission laboratory managed and operated
by National Technology and Engineering Solutions of Sandia, LLC., a wholly
owned subsidiary of Honeywell International, Inc., for the U.S.\  Department of
Energy's National Nuclear Security Administration under contract
DE-NA-0003525.

>>>>

%\title{Model Reduction via \methodName}
%\title{Windowed least--squares reduced-order models for dynamical systems}
\title{Windowed least-squares model reduction for dynamical systems}
%\title{The windowed least-squares framework for model reduction of dynamical systems}

@@

The \methodAcronym\ approach can be viewed as a hybrid space--time method and displays 
commonalities with several related efforts. 
First, Ref.~\cite{bui_thesis} tersely formulated a space--time least-squares ROM
and connected this formulation with optimal control; specifically it mentioned the
optimize-then-discretize vs.\ discretize-then-optimize approaches. This work
did not fully develop this approach, eschewing it 
for 
sequential residual minimization in time (i.e., LSPG). The present work thus
formally develops and extends several of the concepts put forth
in Ref.~\cite{bui_thesis}. Next, Ref.~\cite{constantine_strom} developed a
space--time residual minimization formulation for model interpolation.  The
present work distinguishes itself from Ref.~\cite{constantine_strom} in that
(1) this work also considers trial subspaces characterized by a reduction of
the spatial dimension of the FOM only, and (2) we
minimize the time-continuous FOM residual over arbitrary time
windows. We note that,
similar to the current work, Ref.~\cite{constantine_strom} employs
minimization of the time-continuous FOM residual as a starting point; as such,
this work shares some thematic similarities with
Ref.~\cite{constantine_strom}.  Lastly, Ref.~\cite{choi_stlspg} develops a
space--time extension of the LSPG approach that minimizes the time-discrete
FOM residual over the entire time domain.  The present work distinguishes
itself from Ref.~\cite{choi_stlspg} in that (1) this work minimizes the
time-continuous FOM residual, (2) this work minimizes this residual over
arbitrary time windows, and (3) this work considers trial subspaces
characterized by a reduction of the spatial dimension of the FOM only.

\begin{figure} 
\begin{centering} 
\includegraphics[trim={0.1cm 0cm 1cm 0cm},clip,width=0.99\textwidth]{diagram.pdf} 
\caption{Relationship diagram for the \methodAcronym\ framework for model
	reduction. \KTC{Inconsistent capitalization. I recommend using (1)
	``Limiting case'', (2) Space--time LSPG. Also, can we make this more specific? In particular, can
	we write specific equations (e.g., ``$W = T$ in Eq.~(10)'') for the limiting
	cases?}} 
\label{fig:flowchart} 
\end{centering} 
\end{figure}

 In summary, specific contributions of this work include:
\begin{enumerate}
%\item The development of a time-continuous residual minimization framework for reduced-order models.
\item The \methodNameLower\ (\methodAcronym) framework for dynamical-system
	model reduction. The approach sequentially minimizes the time-continuous
		full-order model residual over arbitrary time windows.
\item Support of two space--time trial subspaces: one that reduces the spatial
	dimension of the FOM only, and one that reduces both the spatial and
		temporal dimensions of the FOM. The former case is of particular
		interest in the \methodAcronym\  contex, as the stationary conditions are
		derived via the Euler--Lagrange equations and comprise a coupled two-point
		Hamiltonian boundary value problem containing a forward and backward
		system. The forward system, which is forced by an auxiliary costate,
		evolves the (spatial) generalized coordinates of the ROM in time. The
		backward system, which is forced by the time-continuous FOM residual
		evaluated about the ROM state, governs the dynamics of the costate. 
	\item Derivation of two solution techniques:
		\textit{discretize-then-optimize} (i.e., direct) and
		\textit{optimize-then-discretize} (i.e., indirect). 
	\item Derivations of conditions under which the
		\methodAcronym\ framework recovers Galerkin, LSPG, and space--time LSPG
		projection.
\item Error analysis of the \methodAcronym\ framework using
	trial subspaces characterized by a reduction of
the spatial dimension of the FOM only.
	This analysis demonstrates that the \methodAcronymROMs\ error 
		is bounded \textit{a priori} by a combination of the
		error at the start of the window and the integrated FOM ODE residual
		evaluated at the FOM state projected onto the trial subspace. As a result,
		the error bound exhibits exponential growth in the number of time windows
		(not total time instances), yielding smaller error-bound growth in thime than
		Galerkin or LSPG projection. \KTC{Check this}
\item Numerical experiments for 
	trial subspaces characterized by a reduction of
the spatial dimension of the FOM only, which demonstrate two key findings:
\begin{itemize}
\item Minimizing the residual over a larger time window leads to more stable
	and physically relevant solutions \KTC{What is `physically relevant'?} as
		well as smaller space--time residual norms. 
\item Minimizing the residual over a larger time window does not necessarily
	lead to a more accurate trajectory (as measured in the space--time
		$\elltwo$-norm of the solution). Instead, minimizing the residual over an
		intermediate-sized time window leads to the smallest trajectory error.
\end{itemize}
\end{enumerate}

The paper proceeds as follows: Section~\ref{sec:math}
outlines the mathematical setting for the full-order model, along with the Galerkin, LSPG, and
space--time LSPG projection. Section~\ref{sec:tclspg} outlines the proposed \methodAcronym\
framework. Section~\ref{sec:numerical_techniques}
outlines numerical techniques for solving \methodAcronymROMs, including both direct and
indirect methods. Section~\ref{sec:analysis} provides
equivalence conditions and error analysis for \methodAcronymROMs.
Section~\ref{sec:numerical_experiments} presents numerical experiments.
Section~\ref{sec:conclude} provides conclusions and perspectives.
We denote vector-valued functions with italicized bold symbols (e.g., $\boldsymbol
x$), vectors with standard bold symbols (e.g., $\mathbf{x}$), 
matrices with capital bold symbols (e.g., $\mathbf{X} \equiv \begin{bmatrix}
\mathbf{x}_1 & \cdots & \mathbf{x}_r\end{bmatrix}$), and spaces with
calligraphic symbols (e.g., $\mathcal{X}$).
%Model reduction of nonlinear dynamical systems is a growing research area.
%The least-squares Petrov-Galerkin (LSPG) approach has gained popularity as a model reduction technique for nonlinear dynamical systems. Derived at the fully discrete level, the LSPG approach solves a least-squares problem that minimizes the fully discrete residual of the reduced-order model at each time-step. LSPG has been shown to be an effective formulation for nonlinear model reduction on complex problems of interest.
%\begin{comment} 
\section{Mathematical formulation}\label{sec:math}
	\KTC{HERE}
	We begin by providing the formulation for the full-order model,
	followed by a description of standard model-reduction methods
	classified according to the type of trial subspace they employ.
	\subsection{Full-order model}\label{sec;FOM}
We consider the system of nonlinear ordinary differential equations (ODEs),
\begin{equation}\label{eq:FOM}
 \stateFOMDotArg{t}  = \velocity(\stateFOMArg{}{t},t), \qquad \stateFOM(0) =
	\stateFOMIC,\qquad t \in [0,T]
\end{equation}
where $\stateFOMDot \equiv \frac{d \stateFOM}{dt}$, 
$T \in \mathbb{R}^+$ denotes the final time, $\stateFOM \in \RR{\fomdim}
	\otimes \timeSpace $ with $\stateFOM: [0,T] \rightarrow  \RR{\fomdim}$
	denotes the state implicitly defined as the solution to initial value
	problem~\eqref{eq:FOM}, 
	$\timeSpace$ denotes the set of real-valued functions acting on the time
	domain, 
 $\stateFOMIC \in \mathbb{R}^{\fomdim}$ denotes the initial condition,
	and $\velocity: \mathbb{R}^{\fomdim} \times [0,T] \rightarrow
	\mathbb{R}^{\fomdim}$ with $(\stateyDiscrete,t) \mapsto
	\velocity(\stateyDiscrete,t)$ denotes the velocity, which is possibly
	nonlinear in its first argument. We refer to Eq.~\eqref{eq:FOM} 
	as the ``full-order model" (FOM) ODE.

@@

\subsection{Reduced-order models}
Projection-based ROMs generate approximate solutions to the FOM
	ODE~\eqref{eq:FOM} by approximating the state in a low-dimensional trial
	subspace. For dynamical 
	systems, two distinct types of space--time trial subspaces are commonly used: 
\begin{enumerate} 
	\item \textit{Space–time trial subspaces with spatial dimension reduction (\spatialAcronym)}. These 
	trial subspaces reduce only the spatial dimension of the FOM, and are commonly used
		in model-reduction methods based on spatial projection followed by time
		integration (e.g., Galerkin projection, LSPG projection.)
	\item \textit{Space–time trial subspaces with spatial and temporal dimension reduction (\spaceTimeAcronym)}. These 
	trial subspaces reduce both the spatial and temporal dimensions of the FOM,
		and are commonly employed 
		employed in model-reduction methods based on space--time projection (e.g.,
		space--time LSPG projection).
\end{enumerate}
 We now describe these two types of trial subspaces and their application to 
	Galerkin, LSPG, and space--time LSPG projection.

@@

\section{Acknowledgments}
The authors thank Yukiko Shimizo and Patrick Blonigan for numerous conversations from which this work benefited. 
E.\ Parish acknowledges an appointment to the Sandia National Laboratories'
John von Neumann Postdoctoral Research Fellowship in Computational Science. 
This work was partially sponsored by Sandia's Advanced Simulation and
Computing (ASC) Verification and Validation (V\&V) Project/Task
\#103723/05.30.02.  This paper describes objective technical results and
analysis. Any subjective views or opinions that might be expressed in the
paper do not necessarily represent the views of the U.S.\ Department of Energy
or the United States Government.  Sandia National Laboratories is a
multimission laboratory managed and operated by National Technology and
Engineering Solutions of Sandia, LLC., a wholly owned subsidiary of Honeywell
International, Inc., for the U.S.\  Department of Energy's National Nuclear
Security Administration under contract DE-NA-0003525.
